---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## 1. Loading and preprocessing the data - Code for reading in the dataset and/or processing the data
```{r load data, echo=TRUE}
data <- read.csv("activity.csv", header = T, sep = ",")
```


## What is mean total number of steps taken per day?

```{r Calulate no of steps per day, echo=TRUE}
sum_of_steps <- tapply(data$steps, data$date, sum, na.rm=T)
```
## 2. Histogram of the total number of steps taken each day
```{r Plot sum of steps per day, echo=TRUE}
hist(sum_of_steps,breaks = 10, col=5,xlab = "sum of steps per day", main = "histogram of steps per day")
```
##3.Mean and median number of steps taken each day 
```{r Find mean and median, echo=TRUE}
mean_sum_of_steps <- mean(sum_of_steps)
median_sum_of_steps <- median(sum_of_steps)
```
```{r Print mean and median, echo=TRUE}
print(c("mean of the total number of steps taken per daymean is:",mean_sum_of_steps, "median of the total number of steps taken per day is:",median_sum_of_steps))
```

## What is the average daily activity pattern?

```{r Calulate mean no of steps for interval, echo=TRUE}
min_int <- tapply(data$steps, data$interval, mean, na.rm=T)
```
##4.Time series plot of the average number of steps taken
```{r Plot Interval Vs Average no of steps, echo=TRUE}
plot(min_int ~ unique(data$interval), col=2,type="l", xlab = "5 mins interval",ylab="Average no of steps")
```
##5. The 5-minute interval that, on average, contains the maximum number of steps
```{r Find interval with max no of steps, echo=TRUE}
max_no_of_steps_int<-min_int[which.max(min_int)]
```
```{r Print interval with max no of steps and Average no of steps, echo=TRUE}
print(c("5 minute interval with max no of steps: ","No of steps:",max_no_of_steps_int))
```
##6. Code to describe and show a strategy for imputing missing data
```{r Find missing values, echo=TRUE}
step_missing_values<-sum(is.na(data$steps))
date_missing_values<-sum(is.na(data$date))
interval_missing_values<-sum(is.na(data$interval))
```
```{r Print Summary of missing values, echo=TRUE}
print(c("step missing values:",step_missing_values,"date missing values:",date_missing_values,"interval missing values:",interval_missing_values))
```


## Imputing missing values

```{r Impute NA values, echo=TRUE}
data1 <- data
for (i in 1:nrow(data)){
    if(is.na(data$steps[i])){
        data1$steps[i]<- min_int[[as.character(data[i, "interval"])]]
    }
}
```
```{r Calulate no of steps per day in imputed data, echo=TRUE}
sum_of_steps1 <- tapply(data1$steps, data1$date, sum, na.rm=T)
```
##7. Histogram of the total number of steps taken each day after missing values are imputed
```{r Plot sum of steps per day in imputed data, echo=TRUE}
hist(sum_of_steps1,breaks = 10, col=5,xlab = "sum of steps per day", main = "histogram of steps per day")
```
```{r Find mean and median in imputed data, echo=TRUE}
mean_sum_of_steps1 <- mean(sum_of_steps1)
median_sum_of_steps1 <- median(sum_of_steps1)
```
```{r Print mean and median imputed data, echo=TRUE}
print(c("mean of the total number of steps taken per daymean is:",mean_sum_of_steps1, "median of the total number of steps taken per day is:",median_sum_of_steps1))
```
Observation
We can observe that both mean and median of imputed data are higher than the mean and median of raw data, suprisingly mean and median for the imputed data is same. 


## Are there differences in activity patterns between weekdays and weekends?
```{r Convert date from factor to date, echo=TRUE} 
data1$date<-as.Date(data1$date)
```
```{r Create a factor variable "day "to store the day of the week, echo=TRUE}
data1$day <- as.factor(weekdays(data1$date))
```
```{r Create logical variable "weekday", echo=TRUE}
data1$weekday <- ifelse(!(data1$day %in% c("Saturday","Sunday")), TRUE, FALSE) 
```

```{r Average number of steps for weekdays, echo=TRUE}
weekdays_data <- data1[data1$weekday,]
steps_per_interval_weekdays <- aggregate(weekdays_data$steps, by=list(interval=weekdays_data$interval), FUN=mean)
```

```{r Average number of steps for weekends, echo=TRUE}
weekends_data <- data1[!data1$weekday,]
steps_per_interval_weekends <- aggregate(weekends_data$steps, by=list(interval=weekends_data$interval), FUN=mean)
```
```{r Add columns names, echo=TRUE}
colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")
colnames(steps_per_interval_weekends) <- c("interval", "average_steps")
#Adding a column to indecate the day
steps_per_interval_weekdays$day <- "Weekday"
steps_per_interval_weekends$day <- "Weekend"
```
```{r Merging the two day types together, echo=TRUE}
week_data <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)
#Converting the day variabke to a factor
week_data$day <- as.factor(week_data$day)
```
##8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r Plotting the data, echo=TRUE}
library(lattice)
xyplot(average_steps ~  interval | day, data = week_data, col=3,layout = c(1,2), type ="l", ylab="Number of Steps")
```
